---
title             : "Reply to: Controlling for non-independence of nations should not be the default choice in cross-cultural research"
shorttitle        : "Non-independence reply"

author: 
  - name          : "Scott Claessens"
    affiliation   : "1*"
  - name          : "Thanos Kyritsis"
    affiliation   : "1"
  - name          : "Quentin D. Atkinson"
    affiliation   : "1,2*"

affiliation:
  - id            : "1"
    institution   : "\\footnotesize School of Psychology, University of Auckland, Auckland, New Zealand"
  - id            : "2"
    institution   : "\\footnotesize School of Anthropology and Museum Ethnography, University of Oxford, Oxford, United Kingdom \\newline *email: scott.claessens@gmail.com; q.atkinson@auckland.ac.nz"

bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
csl               : nature.csl
---
 
\linenumbers

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```

It is widely acknowledged that failing to account for non-independence in data 
can bias statistical estimates and that national-level data often exhibit such 
non-independence. In our recent paper[@Claessens2023], we show that, despite 
this awareness, most analyses of cross-national variation in psychological 
values or economic indicators do not make any attempt to control for 
non-independence, and for those that do, most methods deployed do little to
reduce bias in simulated datasets with non-independence due to proximity or
shared cultural ancestry. We further show that reanalysing a small sample of
published datasets using the best performing methods from our simulations can 
appreciably change parameter estimates.

In a commentary on our paper, Akaliyski[@Akaliyski2024] highlights potential 
issues with our study design, distinguishes between different sources of 
confounding that can arise from non-independence, and proposes causal models 
where controlling for non-independence might actually bias, rather than help, 
inference. For these reasons, Akaliyski argues that controlling for 
non-independence should not be the default analytic choice in cross-national
research.

We appreciate this commentary on our work. It is useful to discuss these 
methodological issues in more detail to improve the rigour of cross-national 
research. That said, we feel that the commentary misinterprets our 
argument in places and overstates potential concerns.

It is worth clarifying from the outset that nowhere in the paper do we argue 
that controlling for non-independence should be the default analytic choice 
in cross-national analyses. Instead of proposing a one-size-fits-all 
solution, we explicitly recommend that "individual studies must outline 
their own particular causal assumptions, which... can then be used to design 
tailored statistical estimation strategies"[@Claessens2023] (p. 8). If controls
for non-independence are likely to bias inference, then they should not be 
included. However, as we will see, we believe these scenarios to be much 
rarer in the real world than Akaliyski claims.

The commentary raises concerns with our particular non-independence controls, 
our simulations, and our reanalyses. First, while it is interesting that 
Akaliyski[@Akaliyski2017] finds a reduced effect of geographic and linguistic 
distances in some regressions, our study and a large body of work 
[@Currie2021;@Diamond1997;@Dow2008;@Gallup1999;@Guglielmino1995;@Kelly2020;
@Kyritsis2022;@Matthews2016] shows that geography and language are important 
explanatory factors for a range of cross-cultural variables. Second, in our 
simulations we generated data with levels of spatial and cultural 
non-independence that were comparable to the strength of non-independence 
found in real-world datasets. We grant this does not prove that the controls 
for non-independence will work in real-world scenarios, but such a proof 
seems an impossibly high bar given that we can never know the true causal 
model in the real world. Third, we acknowledge in the paper that our 
reanalyses do not include all controls from the original studies, meaning
that "we are unable to outright reject the claims from these studies"
[@Claessens2023] (p. 9). Our more modest goal was to show that adding 
controls for non-independence in a stepwise fashion can have an appreciable
impact on reported relationships in real-world data.

Table 1 in the commentary distinguishes between two sources of 
confounding that can arise from non-independence: confounding via diffusion 
(the eponymous Galton's Problem) and confounding via third variables. This 
is a useful contribution, as in hindsight we did not clearly delineate these 
two sources of confounding. In the former scenario, we must control for 
non-independence to account for local and historical diffusion of outcome 
and predictor variables. In the latter scenario, we would like to control 
for confounds such as climate, physical topography, cultural norms, and 
institutions. If these variables are unobserved, then, to the extent that
they are autocorrelated in space or down cultural genealogies, we can 
potentially use geographic and cultural phylogenetic distances to account 
for their influence. Of course, if these third variables are observed, then 
we should just include them directly: nowhere do we propose using controls 
for non-independence as stand-ins for available direct measures. But if we 
allow that unobserved confounds likely exist, then exploring the effect of
including controls for non-independence seems prudent, at the very least.

After defining different sources of confounding, Akaliyski proposes two 
causal models where controlling for non-independence might actually bias 
inference. Akaliyski's Models 4 and 5 illustrate scenarios where the predictor 
variable (X) is non-independent, but the outcome variable (Y) is not. Under 
such models, controlling for non-independence (Z) could either harm precision 
or induce bias. Akaliyski argues that models like these may be more common 
in reality than models where both variables are non-independent (Model 1) 
because processes of diffusion are more likely to influence predictor 
variables than outcome variables. If this is true, controls for 
non-independence may harm inference more often than they will help.

However, compared to Model 1, Models 4 and 5 make much stronger and more 
unrealistic assumptions about the causal relationships between 
national-level variables. These models assume that spatial proximity and 
cultural ancestry influence the outcome variable solely through the 
predictor variable (Z → X → Y) and not through any other causal paths. This 
seems to us an unlikely scenario. It is very likely that outcome variables
will also be affected by spatial proximity and cultural ancestry, either 
directly or indirectly through their effects on unobserved variables.

We can illustrate this by expanding Akaliyski's example of collectivism 
predicting mask wearing, using generative simulations to inform our intuitions 
(Figure \@ref(fig:plot)). If shared ancestry only influences mask 
wearing through collectivism as a mediator, as Akaliyski claims, then 
controlling for shared ancestry does slightly harm the precision of estimates
(Model 4a in Figure \@ref(fig:plot)). But shared ancestry also likely 
influences many other national-level variables that have been shown to be 
related to COVID-19 preventative behaviours, including but not limited to 
cultural tightness[@Gilliam2022], long-term orientation[@Ma2022], uncertainty
avoidance[@Huang2023], religiosity[@Trepanowski2022], and trust in science
and government[@Chen2022]. If we have relevant data on all of these variables, 
then we can control for them directly, but if any are unobserved then it is 
necessary to control for shared ancestry to reduce bias (Model 4b in Figure 
\@ref(fig:plot)).

(ref:plotCaption) Results of simulations testing the implications of controlling for non-independence under different generative causal models. In all simulations, the true causal effect of X on Y is 0. All other causal paths are set to 1. Z represents spatial or cultural phylogenetic non-independence. All other variables are standard normal variables. _N_ = 2000 in each simulated dataset. Densities show the posterior causal effect of X on Y from models fitted without controls for non-independence (red), with controls for non-independence (green), or with a control for non-independence included as an instrumental variable (blue). The IV control is only fitted to data generated from Model 5a, as Z only meets the criteria for an instrumental variable under this generative causal model.

```{r plot, fig.width=6, fig.height=5, fig.cap="(ref:plotCaption)"}
tar_read(plot)
```

Akaliyski then introduces an unobserved variable --- perceived pathogen threat 
--- which influences both collectivism and mask wearing. Indeed, under this 
generative model, controlling for shared ancestry results in bias 
amplification[@Cinelli2022] (Model 5a in Figure \@ref(fig:plot)). 
But again, this causal model is unrealistic. If we allow that perceived pathogen
threat could be influenced by shared ancestry[@Bromham2018] and that shared 
ancestry could influence mask wearing indirectly through any of the unobserved 
variables listed above, then it is necessary to control for shared ancestry to 
reduce bias (though bias still remains due to the unblocked backdoor path 
X ← U~1~ → Y; Model 5b in Figure \@ref(fig:plot)).

Even in situations where Models 4a or 5a do hold, we can still use controls for 
non-independence to estimate unbiased causal effects. For example, in Model 5a,
Z fits the criteria for an instrumental variable: Z is independent of unmeasured
confounds and only influences Y through its effect on X[@McElreath2020]. In this 
situation, including Z as an instrumental variable adjusts for unmeasured
confounding, producing an unbiased estimate of the causal effect (Model 5a in 
Figure \@ref(fig:plot)).

In sum, while Akaliyski's commentary illustrates several theoretical cases 
where controls for non-independence could hinder inferences, we contend that 
such cases are likely to be rare in practice. This is because national-level 
predictor and outcome variables (even contemporary variables) are likely to be 
influenced by a host of factors that are spatially and culturally patterned. 
Even in rare situations where such theoretical cases hold, controls for 
non-independence can still be useful to include if implemented in a manner 
consistent with the underlying causal assumptions (e.g., as instrumental 
variables). Thus, we still think it critically important researchers consider
that cross-national analyses require additional controls to account for the 
non-independence of nations.

\newpage
\nolinenumbers

# Data Availability

All simulated data in this manuscript can be reproduced using the code on 
GitHub: https://github.com/ScottClaessens/crossNationalCorrelationsReply

# Code Availability

All code to reproduce the simulations in this manuscript can be found on 
GitHub: https://github.com/ScottClaessens/crossNationalCorrelationsReply

# Acknowledgements

This work was supported by a Royal Society of New Zealand Marsden grant 
(20-UOA123) to QDA. We thank Erik Ringen for providing feedback on a previous
version of the manuscript and for his valuable input on instrumental variables.

# Author Contributions Statement

SC and QDA wrote the original draft of the manuscript. SC ran the simulations 
and created the figure. All authors reviewed and edited the final draft of the 
manuscript.

# Competing Interests Statement

The authors declare no competing interests.

\newpage

# References

\begingroup
<div id="refs" custom-style="Bibliography"></div>
\endgroup
